---
title: "Analysis of honey bee wing shape in northwestern Europe"
subtitle: "<https://github.com/DrawWing/Apis-mellifera-NW-Europe>"
author: 
  - name: "Julita Machlowska"
    affiliation: University of Agriculture in Krakow, Poland
    orcid: 0000-0001-9951-9248
  - name: "Dariusz Gerula"
    affiliation: National Institute of Horticulture Research, Poland
    orcid: 0000-0003-3875-2657
  - name: "Andrzej Oleksa"
    affiliation: Kazimierz Wielki University in Bydgoszcz, Poland
    orcid: 0000-0002-0414-8075
  - name: "Henk Kok"
    affiliation: Nederlandse Bijenhoudersvereniging, Netherlands
  - name: "Grace P. McCormack"
    affiliation: University of Galway, Ireland
    orcid: 0000-0003-4960-5185
  - name: "Alexandra Valentine"
    affiliation: University of Galway, Ireland
    orcid: 0009-0002-4748-3735
  - name: "Lars Kirkerud"
    affiliation: Norsk Brunbielag, Norway
  - name: "Per Kryger"
    affiliation: Aarhus University, Denmark
    orcid: 0000-0001-5306-3230
  - name: "Laima Blažytė-Čereškienė"
    affiliation: Nature Research Centre, Lithuania
    orcid: 0000-0002-0801-0196
  - name: "Martin Hasselmann"
    affiliation: University of Hohenheim, Germany
    orcid: 0000-0003-3305-4024
  - name: "Teweldemedhn Gebretinsae Hailu"
    affiliation: University of Hohenheim, Germany
    orcid: 0000-0002-3043-995X
  - name: "Benjamin Rutschmann"
    affiliation: Agroscope, Switzerland
    orcid: 0000-0001-6589-6408
  - name: "Adam Tofilski"
    affiliation: University of Agriculture in Krakow, Poland
    orcid: 0000-0002-3898-7029
date: today
date-format: iso
format: 
  html:
    self-contained: true
    toc: true
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	tidy = TRUE, tidy.opts = list(arrow=TRUE, indent=2)
)
setwd("C:/my_programs/R/Apis-mellifera-NW-Europe")

```

## Libraries

```{r libraries}
# calculations
library(geomorph) # GPA 
library(Morpho) # LDA
library(stats) # MANOVA
library(phangorn) # UPGMA
library(IdentiFlyR)  # classification
library(reshape2)  # melt
library(stringr) # str_split_fixed
library(emmeans) # emmeans

# plotting and visualization
library(ggplot2)# plots
ggplot2::theme_set(theme_light())
library(rnaturalearth) # maps
library(terra) # raster maps
library(ggpmisc) # stat_poly_eq
library(ggpubr)  # ggarange
library(ggforce)

```

## Variable names

```{r var-names}
p = 19 # number of landmarks
k = 2  # number of dimensions, in this case 2 for coordinates (x, y)

# create coordinates names used by IdentiFly
nameXy = c("x1", "y1")
for (i in 2:p) {
  nameXy = c(nameXy, paste0("x", i))
  nameXy = c(nameXy, paste0("y", i))
}
nameXy

# The number of principal components used is 2*p-4 = 34, which is equal to the degrees of freedom
namePc <- paste0("PC", 1:(2*p-4))
namePc

# define which landmarks are connected by lines in wireframe graph
linkX <- c(1, 1, 2, 2, 3,  3, 4,  4,  5, 6, 7,  7,  7, 8,  9,  9, 10, 11, 11, 12, 13, 14, 15, 16, 17)
linkY <- c(2, 3, 4, 5, 6, 19, 6, 10, 12, 8, 8, 14, 19, 9, 10, 15, 11, 12, 16, 13, 18, 15, 16, 17, 18)
linkApis <- cbind(linkX, linkY)

```

## Read data

New data collected for this study are available at Zenodo [@machlowska]. There are use also data published during earlier studies [@oleksa2022, @masaquiza2024, @tofilski]

```{r read-data}
dataRaw <- read.csv("https://zenodo.org/records/17612438/files/BY-raw-coordinates.csv")
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/DE-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/ES_517_573-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/FR-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/GB-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/IE-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/LT-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/NL-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/NO-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17612438/files/PL_254_922-raw-coordinates.csv"))

dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/7244070/files/ES-raw-coordinates.csv"))
dataRaw <- rbind(dataRaw, read.csv("https://zenodo.org/records/7244070/files/PL-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/7244070/files/PT-raw-coordinates.csv"))
dataRaw = rbind(dataRaw, read.csv("https://zenodo.org/records/17621032/files/CU-raw-coordinates.csv"))

# Read geographic coordinates
dataGeo <- read.csv("https://zenodo.org/records/17612438/files/BY-data.csv")
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/DE-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/ES_517_573-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/FR-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/GB-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/IE-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/LT-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/NL-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/NO-data.csv"))
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17612438/files/PL_254_922-data.csv"))
geoDataNew = dataGeo

# number of workers per sample in new datasets
nColony = table(geoDataNew$sample)
nColony <- data.frame(nColony, row.names = 1) # move column 1 to row names
ggplot(nColony, aes(x = Freq)) + 
  geom_histogram(binwidth = 1)
summary(nColony$Freq)

dataEU <- read.csv("https://zenodo.org/record/7244070/files/EU-geo-data.csv")
countryEU = substr(dataEU$sample, 1, 2)
dataEU = dataEU[countryEU %in% c("ES", "PL", "PT"),]
dataEU$date = as.numeric(substr(dataEU$date, 1, 4))
dataGeo = rbind(dataGeo, dataEU)
dataGeo = rbind(dataGeo, read.csv("https://zenodo.org/records/17621032/files/CU-data.csv"))

dataGeoSample <- aggregate( dataGeo[c("latitude","longitude","date")], by = list(dataGeo$sample), FUN = mean)
dataGeoSample <- data.frame(dataGeoSample, row.names = 1) # move column 1 to row names

dataRaw$ISO = substr(dataRaw$file, 1, 2)
dataGeoSample$ISO = substr(rownames(dataGeoSample), 1, 2)
dataGeoSample$ISO = as.factor(dataGeoSample$ISO)
dataGeoSample$country = dataGeoSample$ISO
dataGeoSample$country <- gsub("BY", "Belarus", dataGeoSample$country)
dataGeoSample$country <- gsub("CU", "Cuba", dataGeoSample$country)
dataGeoSample$country <- gsub("DE", "Germany", dataGeoSample$country)
dataGeoSample$country <- gsub("ES", "Spain", dataGeoSample$country)
dataGeoSample$country <- gsub("FR", "France", dataGeoSample$country)
dataGeoSample$country <- gsub("GB", "UK", dataGeoSample$country)
dataGeoSample$country <- gsub("IE", "Ireland", dataGeoSample$country)
dataGeoSample$country <- gsub("LT", "Lithuania", dataGeoSample$country)
dataGeoSample$country <- gsub("NL", "Netherlands", dataGeoSample$country)
dataGeoSample$country <- gsub("NO", "Norway", dataGeoSample$country)
dataGeoSample$country <- gsub("PL", "Poland", dataGeoSample$country)
dataGeoSample$country <- gsub("PT", "Portugal", dataGeoSample$country)

# number of samples per location
location = paste(round(dataGeoSample$latitude, 1), round(dataGeoSample$longitude, 1))
nLocation = table(location)
nLocation = as.data.frame(nLocation)

locationLatLon = str_split_fixed(nLocation$location, " ", 2)
colnames(locationLatLon) = c("latitude", "longitude")
locationLatLon = apply(locationLatLon, 2, as.numeric)
locationLatLon = as.data.frame(locationLatLon)
nLocation = cbind(nLocation, locationLatLon)


location = paste(round(dataGeoSample$latitude, 1), round(dataGeoSample$longitude, 1))
nLocation = table(location)
nLocation = as.data.frame(nLocation)

locationLatLon = str_split_fixed(nLocation$location, " ", 2)
colnames(locationLatLon) = c("latitude", "longitude")
locationLatLon = apply(locationLatLon, 2, as.numeric)
locationLatLon = as.data.frame(locationLatLon)
nLocation = cbind(nLocation, locationLatLon)

# average data for country
dataGeoCountry = aggregate( dataGeoSample[c("latitude","longitude")], by = list(dataGeoSample$ISO), FUN = mean)
country = aggregate( dataGeoSample[c("country")], by = list(dataGeoSample$ISO), FUN = unique)
dataGeoCountry = merge(dataGeoCountry, country, by="Group.1")

colnames(dataGeoCountry)[1] = "ISO"
# dataGeoCountry$ISO = unique(dataGeoSample$ISO)
# dataGeoCountry <- data.frame(dataGeoCountry, row.names = 1) # move column 1 to row names
# dataGeoCountry = dataGeoCountry[row.names(dataGeoCountry)!="Cuba",]

```

## Map

```{r map}

# Read elevation data
# In order to download the TIF file uncomment the two lines below #download.file("https://secure-web.cisco.com/15MbNLDNcvKRdwHz0C8BuWIBGjqhNz-VGOCGhfT0ba5LztPXbyrwb6epoX-FTjajgD8ZFDd0GJRkhcWz3V-WwLn7u53iy9KDz55qixlb5sECZRIKC-2tSg0tc_ZY2hq9U0S10DGheC9vVDLKudF18BT9aItv7zRwYlAnC_SQ2ODKtG9EhyJxY4GV6kntF7KVOTyBn0eT9PmiW_EJvCBFchY50GZBD6Ul9wyCVarVhNMTz91dmugWBBEpPb0St3lArbpanzUMmb4xrnZtX1Ve4AW4V9Px-yU6hq3_xT-TnHlbpX-oILMRTq8lJlFRDLUMMzJdYdy_64T_9KQCiP1M9qTPrJjl3rcIo5lZJoHawqz1DUYYKxe-u4oyQK77hLQL4OaxSZXWylQf-rly0woYPMDTW0YixWICqzpMBof0YhX8/https%3A%2F%2Fgeodata.ucdavis.edu%2Fclimate%2Fworldclim%2F2_1%2Fbase%2Fwc2.1_2.5m_elev.zip", "wc2.1_2.5m_elev.zip")
#unzip("wc2.1_2.5m_elev.zip")

# or use your local file
elev <- rast("D:/WorldClim/wc2.1_2.5m_elev.tif")

colorElev <- colorRampPalette(
  c("#f7f7f7", "#d9d9d9", "#bdbdbd", "#636363", "#303030"),
  bias = 2
)

xMin <- -10
xMax <- 40
yMin <- 35
yMax <- 65

# Define extent and crop (terra)
e <- ext(xMin - 5, xMax + 5, yMin - 5, yMax + 5) 
elev <- crop(elev, e)
elevDf <- as.data.frame(elev, xy = TRUE, na.rm = TRUE)
colnames(elevDf) <- c("longitude", "latitude", "altitude")

dataMap <- ne_countries(scale = "medium", returnclass = "sf")

n <- length(unique(dataGeoSample$country))

ggplot(data = dataMap) +
  geom_sf() +
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  geom_jitter(
    data = dataGeoSample,
    aes(x = longitude, y = latitude, color = country, shape = country),
    width = 0.5, height = 0.5
  ) +
  scale_shape_manual(name = "", values = 1:n) +
  scale_color_manual(name = "", values = rainbow(n))

ggplot(data = dataMap) +
  geom_raster(
    data = elevDf,
    aes(x = longitude, y = latitude, fill = altitude)
  ) +
  scale_fill_gradientn(colours = colorElev(100)) +
  geom_sf(fill = NA) +
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  geom_point(
    data = nLocation,
    aes(x = longitude, y = latitude, size = Freq)
  ) +
  scale_size_binned_area(
    name = "colonies",
    limits = c(0, 100),
    breaks = c(0, 5, 10, 20, 40)
  ) +
  theme(legend.position = "right")

# only new samples
geoDataNewSample <- aggregate(
  geoDataNew[c("latitude", "longitude", "date")],
  by = list(geoDataNew$sample),
  FUN = mean
)

geoDataNewSample <- data.frame(geoDataNewSample, row.names = 1) 
geoDataNewSample$ISO <- substr(rownames(geoDataNewSample), 1, 2)

ggplot(data = dataMap) +
  geom_sf() +
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  geom_jitter(
    data = geoDataNewSample,
    aes(x = longitude, y = latitude, color = ISO, shape = ISO),
    width = 0.5, height = 0.5
  ) +
  scale_shape_manual(name = "", values = 1:n) +
  scale_color_manual(name = "", values = rainbow(n))


```

## GPA alignment

```{r GPA}
# Convert from 2D array to 3D array
wing3D <- arrayspecs(dataRaw[ ,nameXy], p, k)
dimnames(wing3D)[[3]] <- dataRaw$file

# Align the coordinates using Generalized Procrustes Analysis 
GPA <- gpagen(wing3D, print.progress = FALSE)
# Convert from 3D array to 2D array
dataAligned <- two.d.array(GPA$coords) 
# rename column names from 1.X, 1.Y to x1, y1 ...
colnames(dataAligned) <- gsub("([0-9]+)\\.([XY])", "\\L\\2\\1",
                                    colnames(dataAligned), perl = TRUE) 

# Average coorinates within samples
dataRaw$sample <- substr(dataRaw$file, 1, 7)
dataSample <- aggregate(dataAligned, by = list(dataRaw$sample), FUN = mean)
dataSample <- data.frame(dataSample, row.names = 1) # move column 1 to row names
dataSample = merge(dataSample, dataGeoSample, by = 0)
dataSample <- data.frame(dataSample, row.names = 1) # move column 1 to row names

```

## Sample size

```{r sample}

nrow(geoDataNew) # number of new wings
geoDataNew$ISO = substr(geoDataNew$file, 1, 2)
table(geoDataNew$ISO)# number of new wings per country
nrow(geoDataNewSample) # number of new samples
table(geoDataNewSample$ISO) # number of new samples per country

nrow(dataRaw) # number of all wings
table(dataRaw$ISO)# number of all wings per country
nrow(dataGeoSample) # number of all samples
table(dataGeoSample$ISO) # number of all samples per country

```

## PCA

```{r PCA}
PCA <- prcomp(dataSample[, nameXy], scale. = TRUE)
PCscores <- as.data.frame(PCA$x[,namePc])
PCscores$country = as.factor(dataSample$country)

# create plot labels
variance.tab <- summary(PCA)$importance
variance <- variance.tab["Proportion of Variance", "PC1"]
variance <- round(100 * variance, 1)
xLabel <- paste0("PC1 (", variance, "%)")
variance <- variance.tab["Proportion of Variance", "PC2"]
variance <- round(100 * variance, 1)
yLabel <- paste0("PC2 (", variance, "%)")

n = length(unique(PCscores$country))
figPca = ggplot(PCscores, aes(x = PC1, y = PC2, shape = country, color = country)) +
  geom_point() +
  scale_shape_manual(name ="", values = c(1:n)) +
  scale_color_manual(name ="", values = rainbow(n)) +
  stat_ellipse() + 
  xlab(xLabel) + ylab(yLabel)
figPca

```

## LDA

```{r lda}
# use equal prior probability for all groups
n.gr <- length(unique(PCscores$country)) # number of groups
lda <- CVA(PCscores[namePc], PCscores$country,
                  rounds = 10000, cv = TRUE,
                  prior = rep(1/n.gr, n.gr))
ldaScores <- as.data.frame(lda$CVscores)
colnames(ldaScores) <- gsub("CV ", "LD", colnames(ldaScores))
# ldaScores <- cbind(sample.dataGeo, ldaScores)
ldaScores$country = PCscores$country
ldaScores$ISO = dataGeoSample$ISO

figLda = ggplot(ldaScores, aes(x = LD1, y = LD2, shape = country, color = country)) +
  geom_point() +
  scale_shape_manual(name ="", values = c(1:n)) +
  scale_color_manual(name ="", values = rainbow(n)) +
  stat_ellipse()
figLda

ggplot(ldaScores, aes(x = LD1, y = LD3, shape = country, color = country)) +
  geom_point() +
  scale_shape_manual(name ="", values = c(1:n)) +
  scale_color_manual(name ="", values = rainbow(n)) +
  stat_ellipse()

ggarrange(figPca, figLda, 
          labels = c("a", "b"), 
          font.label = list(size = 12, face = "bold"),  
          ncol = 2, nrow = 1, 
          common.legend = TRUE, legend = "bottom") 

# Confusion matrix
# Classification of the dataSample to groups
ldaclass <- typprobClass(lda$CVscores, groups = ldaScores$ISO, outlier = 0)
print(ldaclass)

# Mahalanobis distances between groups
knitr::kable(as.data.frame(as.matrix(lda$Dist$GroupdistMaha)), digits = 3)
# pairwise probabilities of differences between groups
knitr::kable(as.data.frame(as.matrix(lda$Dist$probsMaha)), digits = 5)

# wing shape differs significantely between countries
MANOVA = manova(as.matrix(cbind(PCscores[, namePc])) ~ country, PCscores)
summary(MANOVA)

```

## Classification as lineages

Comparison with lineages dataset <a href="#Nawrocka et al., 2018a">Nawrocka et al., (2018a)</a> and <a href="#Nawrocka et al., 2018b">Nawrocka et al., (2018b)</a>

```{r colony-lin}

idData <- xml2gmLdaData("https://zenodo.org/records/14054009/files/apis-mellifera-lineage.dw.xml")
idM = gmLdaData2id(idData,
                  dataSample[,nameXy],
                  average = FALSE)
idM$plot
tab = table(idM$id$group)
tab
prop.table(tab) # proportion of colonies classified as each of the lineages

tab = table(idM$id$group, dataSample$ISO)
tab
round(prop.table(tab, margin = 2), digits = 3)

pTab = as.data.frame(idM$P)
sums = rowSums(pTab)
pTab = pTab/sums

pTabLong <- pTab
pTabLong$sample <- rownames(pTabLong)
pTabLong <- melt(pTabLong, id.vars = c("sample"))
pTabLong$gr <- substr(pTabLong$sample, 1, 2)  # short 2 letter name

ggplot(pTabLong, aes(sample, value, fill = variable)) +
  geom_col(width = 1) +
  scale_fill_manual(name="classified as lineage", values = rainbow(4)) +
  facet_grid(~gr, switch = "x", scales = "free", space = "free") +
  theme_minimal() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(expand = c(0, 0)) +
  guides(fill=guide_legend(nrow=1, byrow=TRUE)) +
  theme(panel.spacing.x = unit(0.2, "lines"), axis.text.x = element_blank(),
        legend.position = "top") 

# countrySmallN = c("BY", "CU", "DE", "FR", "GB")
countrySmallN = c("Belarus", "Cuba", "Germany", "France", "UK")
# countries with large N
pTabLong <- pTab
pTabLong = pTabLong[!dataSample$country %in% countrySmallN, ]
pTabLong$sample <- rownames(pTabLong)
pTabLong <- melt(pTabLong, id.vars = c("sample"))
pTabLong$gr <- substr(pTabLong$sample, 1, 2)  # short 2 letter name
figAdmA = ggplot(pTabLong, aes(sample, value, fill = variable)) +
  geom_col(width = 1) +
  scale_fill_manual(name="classified as lineage", values = rainbow(4)) +
  facet_grid(~gr, switch = "x", scales = "free", space = "free") +
  theme_minimal() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(expand = c(0, 0)) +
  guides(fill=guide_legend(nrow=1, byrow=TRUE)) +
  theme(panel.spacing.x = unit(0.2, "lines"), axis.text.x = element_blank(),
        legend.position = "top") 

# countries with small N
pTabLong <- pTab
pTabLong = pTabLong[dataSample$country %in% countrySmallN, ]
pTabLong$sample <- rownames(pTabLong)
pTabLong <- melt(pTabLong, id.vars = c("sample"))
pTabLong$gr <- substr(pTabLong$sample, 1, 2)  # short 2 letter name
figAdmB = ggplot(pTabLong, aes(sample, value, fill = variable)) +
  geom_col(width = 1) +
  scale_fill_manual(name="classified as lineage", values = rainbow(4)) +
  facet_grid(~gr, switch = "x", scales = "free", space = "free") +
  theme_minimal() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(expand = c(0, 0)) +
  guides(fill=guide_legend(nrow=1, byrow=TRUE)) +
  theme(panel.spacing.x = unit(0.2, "lines"), axis.text.x = element_blank(),
        legend.position = "top") + 
  theme(
    axis.text.y  = element_blank(),
  )

figAdm = ggarrange(figAdmA, figAdmB, 
          ncol = 2, nrow = 1, 
          widths = c(0.66, 0.33),
          common.legend = TRUE, legend = "top")

dataGeoSample$lineage  = idM$id$group

ggplot(data = dataMap) + geom_sf() + 
  geom_jitter(data = dataGeoSample, aes(x = longitude, y = latitude, color = lineage), width = 0.4, height = 0.4, size = 0.7) + 
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  scale_color_manual(name = "", values = rainbow(4))

tabProp = prop.table(tab, margin = 2)
tabProp = t(tabProp)
tabProp = tabProp[-2,]
countryLin = as.data.frame(tabProp)
colnames(countryLin) <- c("ISO", "lineage", "Freq")
countryLin = merge(countryLin, dataGeoCountry, by="ISO")

ggplot(countryLin) + 
  geom_arc_bar(aes(x0 = longitude, y0 = latitude, r0 = 0, r = 1.5, amount = Freq, fill = lineage), stat = "pie") + 
  geom_text(aes(x = longitude, y = latitude, label = country)) +  
  scale_fill_manual(name = "", values = rainbow(4)) +
  coord_cartesian(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  coord_fixed() 

figMapClass = ggplot(data = dataMap) + geom_sf() + 
  geom_arc_bar(data = countryLin, aes(x0 = longitude, y0 = latitude, r0 = 0, r = 1, amount = Freq, fill = lineage), stat = "pie")  +
  scale_fill_manual(name = "", values = rainbow(4)) +
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) 
figMapClass

ggarrange(figAdm, figMapClass,
          labels = c("a", "b"), 
          font.label = list(size = 12, face = "bold"),
          ncol = 1, nrow = 2,
          heights = c(0.5, 1.5))

```

## Changes in time

```{r time}

dataId = cbind(dataSample[, c("latitude", "longitude", "date", "country")], idM$MD2, idM$LD)
dataId$sample = row.names(dataSample)
dataPl = dataId[dataId$country == "Poland", ]
tmp = str_split_fixed(dataPl$sample, "-", 2)
tmp = as.numeric(tmp[,2])
dataPl$sampleNr = tmp

dataPl$years = ifelse(dataPl$date<2016,"before 2016", "after 2016")
dataPl$years = factor(dataPl$years, levels = c("before 2016", "after 2016"))
dataPl$population = ifelse((dataPl$sampleNr > 253 & dataPl$sampleNr < 680),
                           "protected", "unprotected")
dataPl$population = as.factor(dataPl$population)
table(dataPl$population, dataPl$years)

fit <- lm(M ~ years*population, data=dataPl)
anova(fit)

emm_date_by_pop <- emmeans(fit, ~ years | population)
pairs(emm_date_by_pop, adjust = "tukey")   

emm_pop_by_date <- emmeans(fit, ~ population | years)
pairs(emm_pop_by_date, adjust = "tukey")

ggplot(dataPl, aes(x=population, y=M)) + 
  geom_boxplot(aes(fill=years)) +
    scale_fill_brewer(palette="BuPu") +
  labs(y ="distance to lineage M")

```

## Information about session

```{r info}

sessionInfo()

```
